---
title: "Water Pumps"

output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## About the Dataset
Using data from Taarifa and the Tanzanian Ministry of Water, we set out to predict what could impact the state of water pumps. 
A smart understanding of which waterpoints will fail can improve maintenance operations and ensure that clean, potable water is available to communities across Tanzania.
More information about the challenge and the dataset can be found here - https://www.drivendata.org/competitions/7/pump-it-up-data-mining-the-water-table/page/23/

**MORE METADATA TO FOLLOW**


## Import Data

```{r}
# Define train_values_url
train_values_url <- "http://s3.amazonaws.com/drivendata/data/7/public/4910797b-ee55-40a7-8668-10efd5c1b960.csv"
WaterPumps_value = read.csv(train_values_url, header=TRUE, stringsAsFactors = FALSE)
# Define train_labels_url
train_labels_url <- "http://s3.amazonaws.com/drivendata/data/7/public/0bf8bc6e-30d0-4c50-956a-603fc693d966.csv"
WaterPumps_label = read.csv(train_labels_url, header=TRUE, stringsAsFactors = TRUE)

WaterPumps = merge(WaterPumps_values, WaterPumps_labels, by='id')

# Define test_values_url
test_values_url <- "http://s3.amazonaws.com/drivendata/data/7/public/702ddfc5-68cd-4d1d-a0de-f5f566f76d91.csv"
```

## Data Cleaning:

```{r}
#Create New Variables:

```

Steven Was Here
```{r}
# Clean Data

```


## Exploratory Analysis:

```{r}

```

## Variable Selection

```{r}

```

## Modeling

```{r}

```

## Evaluation & Selection

```{r}

```


